---
title: "CRAN Download Analysis for {echos}"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    df_print: paged
    theme:
      bg: "white"
      fg: "#202123"
      primary: "steelblue"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)

library(adjustedcranlogs)
library(dplyr)
library(lubridate)
library(ggplot2)
library(scales)
library(zoo)
library(patchwork)
library(tscv)
library(gt)
library(glue)

# Change default location and time
Sys.setlocale("LC_TIME", "C")
```

```{css, echo=FALSE}
h1.title {
  font-weight: 700;
}

h1, h2, h3 {
  font-weight: 700;
}
```

```{r configuration, include=FALSE}
package_name <- "echos"

start_date <- as.Date("2025-02-01")
end_date   <- Sys.Date()

rolling_window <- 7
release_window_days <- 30

col_main  <- "steelblue"
col_light <- "lightsteelblue"

release_dates <- tibble::tribble(
  ~version, ~release_date,
  "v1.0.1", as.Date("2025-02-11"),
  "v1.0.2", as.Date("2025-06-23"),
  "v1.0.3", as.Date("2026-02-22")
)
```

```{r download-data, include=FALSE}
downloads_raw <- adj_cran_downloads(
  packages = package_name,
  from = start_date,
  to = end_date
)

downloads <- downloads_raw |>
  mutate(
    date = as.Date(date),
    month = floor_date(date, unit = "month"),
    downloads = pmax(adjusted_downloads, 0)
  )
```

```{r calculations, include=FALSE}
downloads <- downloads |>
  arrange(date) |>
  mutate(
    rolling_downloads = rollmean(
      downloads,
      k = rolling_window,
      fill = NA,
      align = "right"
    ),
    cumulative_downloads = cumsum(downloads)
  )

analysis_start_date <- min(downloads$date, na.rm = TRUE)
analysis_end_date <- max(downloads$date, na.rm = TRUE)

total_downloads <- sum(downloads$downloads, na.rm = TRUE)
mean_daily_downloads <- mean(downloads$downloads, na.rm = TRUE)
median_daily_downloads <- median(downloads$downloads, na.rm = TRUE)
sd_daily_downloads <- sd(downloads$downloads, na.rm = TRUE)
min_daily_downloads <- min(downloads$downloads, na.rm = TRUE)
max_daily_downloads <- max(downloads$downloads, na.rm = TRUE)

last_cumulative_observation <- downloads |>
  slice_tail(n = 1)

daily_download_summary <- tibble::tibble(
  Statistic = c(
    "Minimum",
    "Mean",
    "Median",
    "Standard deviation",
    "Maximum",
    "Total"
  ),
  Value = c(
    min_daily_downloads,
    mean_daily_downloads,
    median_daily_downloads,
    sd_daily_downloads,
    max_daily_downloads,
    total_downloads
  )
)

monthly_downloads <- downloads |>
  group_by(month) |>
  summarise(
    downloads = sum(downloads, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(month) |>
  mutate(
    downloads_lag = lag(downloads),
    change_abs = downloads - downloads_lag,
    change_pct = downloads / downloads_lag - 1,
    label = if_else(
      is.na(change_pct),
      comma(downloads),
      paste0(
        comma(downloads),
        "\n",
        if_else(change_pct > 0, "+", ""),
        percent(change_pct, accuracy = 1)
      )
    )
  )

last_observed_month <- floor_date(analysis_end_date, unit = "month")
last_day_observed_month <- ceiling_date(last_observed_month, unit = "month") - days(1)

complete_monthly_downloads <- if (analysis_end_date < last_day_observed_month) {
  monthly_downloads |>
    filter(month < last_observed_month)
} else {
  monthly_downloads
}

strongest_month <- complete_monthly_downloads |>
  slice_max(downloads, n = 1, with_ties = FALSE)

latest_month <- monthly_downloads |>
  slice_tail(n = 1)

latest_month_is_incomplete <- analysis_end_date < last_day_observed_month

top_growth_months <- complete_monthly_downloads |>
  filter(!is.na(change_pct), change_pct > 0) |>
  arrange(desc(change_pct)) |>
  slice_head(n = 3) |>
  mutate(
    growth_label = paste0(
      format(month, "%B %Y"),
      " (",
      percent(change_pct, accuracy = 0.1),
      ")"
    )
  )

top_decline_months <- complete_monthly_downloads |>
  filter(!is.na(change_pct), change_pct < 0) |>
  arrange(change_pct) |>
  slice_head(n = 3) |>
  mutate(
    decline_label = paste0(
      format(month, "%B %Y"),
      " (",
      percent(change_pct, accuracy = 0.1),
      ")"
    )
  )

growth_months_text <- if (nrow(top_growth_months) > 0) {
  paste(top_growth_months$growth_label, collapse = ", ")
} else {
  "no months with positive month-over-month growth"
}

decline_months_text <- if (nrow(top_decline_months) > 0) {
  paste(top_decline_months$decline_label, collapse = ", ")
} else {
  "no months with negative month-over-month change"
}

release_points <- release_dates |>
  filter(
    release_date >= min(downloads$date),
    release_date <= max(downloads$date)
  ) |>
  left_join(
    downloads |>
      select(date, cumulative_downloads),
    by = c("release_date" = "date")
  ) |>
  mutate(
    release_label = paste0(version, "\n", format(release_date, "%Y-%m-%d"))
  )

release_impact <- release_dates |>
  rowwise() |>
  mutate(
    downloads_before = sum(
      downloads$downloads[
        downloads$date >= release_date - days(release_window_days) &
          downloads$date < release_date
      ],
      na.rm = TRUE
    ),
    downloads_after = sum(
      downloads$downloads[
        downloads$date > release_date &
          downloads$date <= release_date + days(release_window_days)
      ],
      na.rm = TRUE
    ),
    change_abs = downloads_after - downloads_before,
    change_pct = if_else(
      downloads_before > 0,
      change_abs / downloads_before,
      NA_real_
    )
  ) |>
  ungroup()

release_impact_with_change <- release_impact |>
  filter(!is.na(change_pct))

strongest_release <- release_impact_with_change |>
  slice_max(change_pct, n = 1, with_ties = FALSE)

release_summary_sentences <- release_impact |>
  mutate(
    sentence = if_else(
      is.na(change_pct),
      glue(
        "After **{version}**, downloads reached **{comma(downloads_after)}** in the ",
        "{release_window_days} days after the release. A relative change is ",
        "not reported because there were no downloads in the comparison window before the release."
      ),
      glue(
        "After **{version}**, downloads increased from **{comma(downloads_before)}** ",
        "before the release to **{comma(downloads_after)}** after the release. ",
        "This corresponds to an increase of **{comma(change_abs)} downloads**, ",
        "or **{percent(change_pct, accuracy = 0.1)}**."
      )
    )
  ) |>
  pull(sentence)

release_summary_text <- paste0(
  "- ",
  release_summary_sentences,
  collapse = "\n"
)

daily_label_y <- max(downloads$downloads, na.rm = TRUE) * 0.95
cumulative_label_nudge_y <- max(downloads$cumulative_downloads, na.rm = TRUE) * 0.04
total_label_nudge_y <- -max(downloads$cumulative_downloads, na.rm = TRUE) * 0.01 # 0.05
```

```{r tables, include=FALSE}
daily_download_summary_gt <- daily_download_summary |>
  gt() |>
  tab_header(
    title = "Daily download summary",
    subtitle = paste0(
      "Adjusted downloads, ",
      format(analysis_start_date, "%Y-%m-%d"),
      " to ",
      format(analysis_end_date, "%Y-%m-%d")
    )
  ) |>
  fmt_number(
    columns = Value,
    decimals = 1,
    use_seps = TRUE
  ) |>
  cols_align(
    align = "left",
    columns = Statistic
  ) |>
  cols_align(
    align = "right",
    columns = Value
  ) |>
  tab_options(
    table.font.size = px(12),
    heading.title.font.size = px(14),
    heading.subtitle.font.size = px(11),
    table.background.color = "white",
    heading.background.color = "white",
    column_labels.background.color = "white",
    table.font.color = "#202123"
  )

release_impact_gt <- release_impact |>
  transmute(
    Version = version,
    `Release date` = release_date,
    `Downloads before` = downloads_before,
    `Downloads after` = downloads_after,
    `Absolute change` = change_abs,
    `Relative change` = change_pct
  ) |>
  gt() |>
  tab_header(
    title = "Release impact",
    subtitle = paste0("Adjusted downloads, ±", release_window_days, " days")
  ) |>
  fmt_date(
    columns = `Release date`,
    date_style = "iso"
  ) |>
  fmt_number(
    columns = c(`Downloads before`, `Downloads after`, `Absolute change`),
    decimals = 0,
    use_seps = TRUE
  ) |>
  fmt_percent(
    columns = `Relative change`,
    decimals = 1
  ) |>
  sub_missing(
    columns = everything(),
    missing_text = "—"
  ) |>
  cols_align(
    align = "right",
    columns = c(
      `Downloads before`,
      `Downloads after`,
      `Absolute change`,
      `Relative change`
    )
  ) |>
  tab_options(
    table.font.size = px(12),
    heading.title.font.size = px(14),
    heading.subtitle.font.size = px(11),
    table.background.color = "white",
    heading.background.color = "white",
    column_labels.background.color = "white",
    table.font.color = "#202123"
  )
```

```{r plots, include=FALSE}
p_daily <- ggplot(downloads, aes(x = date))

p_daily <- p_daily +
  geom_line(
    aes(y = downloads),
    color = col_light,
    linewidth = 0.75
  )

p_daily <- p_daily +
  geom_line(
    aes(y = rolling_downloads),
    color = col_main,
    linewidth = 1.5
  )

p_daily <- p_daily +
  geom_vline(
    data = release_points,
    aes(xintercept = release_date),
    linetype = "dotted",
    color = "grey40",
    linewidth = 0.75
  )

p_daily <- p_daily +
  geom_label(
    data = release_points,
    aes(
      x = release_date,
      y = daily_label_y,
      label = release_label
    ),
    size = 3.1,
    vjust = 1,
    label.padding = unit(0.15, "lines")
  )

p_daily <- p_daily +
  scale_y_continuous(
    labels = comma,
    expand = expansion(mult = c(0.02, 0.12))
  )

p_daily <- p_daily +
  labs(
    title = "(a) Daily downloads",
    subtitle = "Adjusted values and 7-day average",
    x = NULL,
    y = NULL
  )

p_daily <- p_daily +
  theme_tscv()

p_cumulative <- ggplot(downloads, aes(x = date, y = cumulative_downloads))

p_cumulative <- p_cumulative +
  geom_line(
    color = col_main,
    linewidth = 1.5
  )

p_cumulative <- p_cumulative +
  geom_vline(
    data = release_points,
    aes(xintercept = release_date),
    linetype = "dotted",
    color = "grey40",
    linewidth = 0.75
  )

p_cumulative <- p_cumulative +
  geom_point(
    data = release_points,
    aes(x = release_date, y = cumulative_downloads),
    color = "grey30",
    size = 2
  )

p_cumulative <- p_cumulative +
  geom_label(
    data = release_points,
    aes(
      x = release_date,
      y = cumulative_downloads,
      label = release_label
    ),
    nudge_y = cumulative_label_nudge_y,
    size = 3.1,
    label.padding = unit(0.15, "lines")
  )

p_cumulative <- p_cumulative +
  geom_point(
    data = last_cumulative_observation,
    color = col_main,
    size = 4
  )

p_cumulative <- p_cumulative +
  geom_label(
    data = last_cumulative_observation,
    aes(label = paste0("Total: ", comma(cumulative_downloads))),
    nudge_x = -20,
    nudge_y = total_label_nudge_y,
    hjust = 1,
    vjust = 1,
    label.padding = unit(0.2, "lines")
  )

p_cumulative <- p_cumulative +
  scale_y_continuous(
    labels = comma,
    expand = expansion(mult = c(0.02, 0.15))
  )

p_cumulative <- p_cumulative +
  scale_x_date(
    expand = expansion(mult = c(0.02, 0.08))
  )

p_cumulative <- p_cumulative +
  labs(
    title = "(b) Cumulative downloads",
    subtitle = "Releases and final total",
    x = NULL,
    y = NULL
  )

p_cumulative <- p_cumulative +
  theme_tscv()

p_daily_histogram <- ggplot(downloads, aes(x = downloads))

p_daily_histogram <- p_daily_histogram +
  geom_histogram(
    bins = 30,
    fill = col_light,
    color = "white"
  )

p_daily_histogram <- p_daily_histogram +
  geom_vline(
    xintercept = mean_daily_downloads,
    color = col_main,
    linewidth = 1,
    linetype = "dashed"
  )

p_daily_histogram <- p_daily_histogram +
  geom_vline(
    xintercept = median_daily_downloads,
    color = "black",
    linewidth = 1,
    linetype = "dotted"
  )

p_daily_histogram <- p_daily_histogram +
  annotate(
    "label",
    x = mean_daily_downloads,
    y = Inf,
    label = paste0("Mean: ", comma(round(mean_daily_downloads, 1))),
    vjust = 1.5,
    hjust = -0.05,
    color = col_main
  )

p_daily_histogram <- p_daily_histogram +
  annotate(
    "label",
    x = median_daily_downloads,
    y = Inf,
    label = paste0("Median: ", comma(median_daily_downloads)),
    vjust = 3.2,
    hjust = -0.05,
    color = "black"
  )

p_daily_histogram <- p_daily_histogram +
  scale_x_continuous(labels = comma)

p_daily_histogram <- p_daily_histogram +
  scale_y_continuous(labels = comma)

p_daily_histogram <- p_daily_histogram +
  labs(
    title = "(c) Distribution",
    subtitle = "Adjusted downloads",
    x = "Adjusted daily downloads",
    y = "Number of days"
  )

p_daily_histogram <- p_daily_histogram +
  theme_tscv()

p_monthly <- ggplot(monthly_downloads, aes(x = month, y = downloads))

p_monthly <- p_monthly +
  geom_col(fill = col_light)

p_monthly <- p_monthly +
  geom_text(
    aes(label = label),
    vjust = -0.3,
    size = 3.2,
    lineheight = 0.9
  )

p_monthly <- p_monthly +
  scale_y_continuous(
    labels = comma,
    expand = expansion(mult = c(0, 0.18))
  )

p_monthly <- p_monthly +
  labs(
    title = "Monthly downloads",
    subtitle = "Adjusted totals and change",
    x = NULL,
    y = NULL
  )

p_monthly <- p_monthly +
  theme_tscv()

core_download_plots <- p_daily /
  (p_cumulative | p_daily_histogram) +
  plot_layout(heights = c(1, 0.9))
```

# Overview

This report summarizes the CRAN download statistics for the R package `{echos}` from **`r format(analysis_start_date, "%Y-%m-%d")` to `r format(analysis_end_date, "%Y-%m-%d")`**.

The analysis uses **adjusted CRAN download data**. These adjusted data are preferable to unadjusted base download counts because raw CRAN downloads can include automated activity, repeated downloads, and other technical effects that do not necessarily represent genuine user interest. Adjusted downloads therefore provide a more conservative and meaningful estimate of actual package usage.

For `{echos}`, the difference between base and adjusted downloads is minimal. Nevertheless, using adjusted data makes the interpretation cleaner and more defensible.

Across the observed period, `{echos}` received **`r comma(total_downloads)` adjusted CRAN downloads**. The package averaged **`r round(mean_daily_downloads, 1)` downloads per day**, with a median of **`r median_daily_downloads` downloads per day**. This suggests a modest but stable level of usage, with additional peaks around release periods.

# Daily downloads

```{r print-core-download-plots, fig.width=14, fig.height=9}
core_download_plots
```

The daily time series shows a recurring baseline of downloads with several visible short-term peaks. The `r rolling_window`-day rolling average smooths out day-to-day variation and makes the underlying trend easier to see. Overall, the package appears to receive regular attention rather than only isolated downloads.

The cumulative plot shows steady growth over time. The slope becomes steeper during periods of stronger download activity, especially around release phases. This indicates that releases and related attention contribute meaningfully to the overall download trajectory.

The distribution plot confirms that most days are close to the regular baseline, while a smaller number of high-download days create the upper tail of the distribution.

```{r print-daily-summary-table}
daily_download_summary_gt
```

The daily summary shows that `{echos}` has developed a stable download baseline. The **mean of `r round(mean_daily_downloads, 1)` downloads per day** and the **median of `r median_daily_downloads` downloads per day** are close to each other, which indicates that the average is not driven only by a small number of extreme days.

The standard deviation of **`r round(sd_daily_downloads, 1)` downloads** shows that daily downloads vary around this baseline, with occasional spikes. The **maximum of `r max_daily_downloads` downloads** shows that the package occasionally receives concentrated attention. These peaks likely reflect release activity, communication around the package, documentation updates, examples, or external references.

The **minimum of `r min_daily_downloads` downloads** is not unusual for a specialized R package. More important is the overall total of **`r comma(total_downloads)` adjusted downloads**, which indicates sustained interest over the full observation period.

# Monthly downloads

```{r print-monthly-plot, fig.width=14, fig.height=5}
p_monthly
```

The monthly download chart shows a pattern of recurring activity rather than a one-time spike. The strongest complete month was **`r format(strongest_month$month, "%B %Y")`**, with **`r comma(strongest_month$downloads)` adjusted downloads**. This suggests that `{echos}` gained noticeable attention during that period.

The months with the strongest positive month-over-month growth were **`r growth_months_text`**. These increases suggest that `{echos}` gained attention at multiple points during the observation period.

The months with the strongest declines were **`r decline_months_text`**. This is not necessarily concerning. For a specialized package, month-to-month variation is expected, especially when downloads are influenced by releases, examples, teaching material, workshops, or short periods of external attention.

`r if (latest_month_is_incomplete) paste0("The value for **", format(latest_month$month, "%B %Y"), "** should be interpreted carefully because the observation period ends on **", format(analysis_end_date, "%Y-%m-%d"), "**. Therefore, this month is not a full-month result.") else paste0("The latest observed month, **", format(latest_month$month, "%B %Y"), "**, is included as a complete month in the current observation period.")`

# Release impact

```{r print-release-impact-table}
release_impact_gt
```

The release impact analysis compares downloads in the **`r release_window_days` days before** and **`r release_window_days` days after** each release. This gives a simple indication of whether releases were followed by higher download activity.

```{r release-impact-text, results='asis'}
cat(release_summary_text)
```

These results suggest that releases are meaningful visibility events for `{echos}`. `r if (nrow(strongest_release) > 0) paste0("Among releases with a valid before/after comparison, **", strongest_release$version, "** showed the strongest relative post-release increase, with downloads rising by **", percent(strongest_release$change_pct, accuracy = 0.1), "**.") else "The first observed release mainly reflects early package uptake, because no before-release comparison window was available."`

# Conclusion

Between **`r format(analysis_start_date, "%Y-%m-%d")` and `r format(analysis_end_date, "%Y-%m-%d")`**, `{echos}` received **`r comma(total_downloads)` adjusted CRAN downloads**. The package reached an average of **`r round(mean_daily_downloads, 1)` adjusted downloads per day**, with visible increases after major releases.

Overall, the results suggest that `{echos}` is gaining traction. The package shows steady baseline use, release-related increases, and recurring periods of stronger download activity.
